Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
finished

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}
\newcommand{\id}{\mathsf{id}}

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Our notation on paper is close
to Agda's, we write $\Pi$ types as $(x:A)\ra B$, $\Sigma$ types as
$(x:A)\times B$, and use implicit arguments and overloading
extensively.

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])\,(\Gamma\ext A)$. So $p'$ has to be a
new kind of weakening which adds a new variable in the last but one
position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\begin{alignat*}{10}
  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\end{alignat*}
where $+^n$ means the $n$-times iteration of $\blank^+$.

@@

@@

    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\

@@

    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\

@@

  A))\,(F\,\Gamma\ext F\,A)$ is an isomorphism.

@@

Starting with an SSC-model, termifying it and comparing the result
with the original SSC does not provide an isomorphism because we can't
define the map $F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on substitutions is
not definable using single substitutions. We leave formulating the
right notion of contextual isomorphism for SSCs as future work.

@@

Why are we still looking at alternative ways of defining type theory?
Closer to the syntax/implementation, easier to understand/implement,
fewer equations, hence easier to define models, coherent syntax?

We showed a particular example for the SOGAT->single GAT translation,
we showed that the syntax is equivalent to the other one, we
conjecture that the method presented here generalises. For this
particular SOGAT, the single GAT syntax can be simplified, and in
fact, it is ``equivalent'' to the parallel theory.

Can we remove even more structural rules? the termification doesn't
work if we don't have e.g. $[\p][^+]$ and $[\p][\langle\rangle]$ for all terms, and

Separate sort for variables, without the extra type equations, by
using assumptions, and proving everything by induction. If we have
var, then we cannot do termification to get a CwF from a var-SSC, see
the proof of equation (\ref{eq:functor}).

Prove full equivalence of CwF+ with SSC+.

>>>>

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Tel}{\mathsf{Tel}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\Type}{\mathsf{Type}}

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda, we use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
overloading extensively.

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
new kind of weakening which adds a new variable in the last but one
position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\begin{alignat*}{10}
  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\end{alignat*}
where $+^n$ means the $n$-times iteration of $\blank^+$.

@@

\subsection{Standard model}
\label{sec:standard}

@@

    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\[-1.5em]

@@

    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\[-1em]

@@

  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack

@@

The other roundtrip, that is, starting with an SSC-model, termifying
it and comparing the result with the original SSC does not provide an
isomorphism because we can't define the map $F :
\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
substitutions. Such a map is not definable only using single
substitutions. However, the set of terms and types are still
isomorphic after the roundtrip. We leave formulating the right notion
of contextual isomorphism for SSCs as future work.

@@

Why are we still looking at alternative ways of defining type theory?
New definitions might be closer to actual implementations, suggest new
implementation techniques, they might be better pedagogically, have
more models, or contribute to a coherent definition. We elaborate on
the latter.

Defining a useable syntax of type theory internally to homotopy type
theory (HoTT \cite{HoTTbook}) is an open problem: the (h)set-truncated
syntax can be formalised as a QIIT, but then we cannot even define the
$\Type$-interpretation, that is, a function from the syntax which
interprets closed types by the metatheoretic universe $\Type$ (rather
than an inductive recursive universe as in Subsection
\ref{sec:standard} or \cite[Section
  4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
syntax of type theory can be defined using infinitely many coherene
rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
$\Sub$, we have the associativity of composition of substitutions
equation. If we truncate our syntax to be a set, then this is
enough. If we don't truncate, there are two different ways of proving
$((\gamma\circ\delta)\circ\theta)\circ\xi =
\gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
be the same, so in such a (wild) syntax substitutions don't form a
set, hence they don't have decidable equality \cite{hedberg}. The
solution is to add the pentagon law, but then this introduces new
higher equalities, which require more coherence laws, so in the end we
have an infinite sequence of higher coherence laws that need to be
part of the syntax. Our single substitution calculus does not have
composition for $\Sub$, so there is no need for higher coherence laws
like associativity. However as it currently stands, our syntax is not
coherent, it needs set-truncation when working in HoTT. For example,
there are two different ways to prove the following equation:
\begin{alignat*}{10}
  & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
  & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
\end{alignat*}
In a coherent version of our syntax we either need to state that these
two proofs are equal, or we have to remove one of them. We can remove
the first one by limiting $[\p][\langle\rangle]$ to only work on
variables (this needs the introduction of a new sort for
variables). If we do this, the minimisation of Section
\ref{sec:minimisation} does not work anymore, so we need the extra
four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
$[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
types make the syntax non-coherent again. A possible way out of this
would be to make the equations that depend on one of these for
equations conditional. But then the methods of Section
\ref{sec:admissible} don't work anymore for re-proving the needed
admissible equations. Also, if only have the rule $[\p][^+]$ only for
variables, then we cannot redo the termification of Section
\ref{sec:cwf}, in particular it is not clear how to prove equation
(\ref{eq:functor}).

In this paper we unfolded a particular example of the SOGAT $\ra$
single substitution GAT translation
\cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
syntax is isomorphic to the parallel syntax, and we conjecture that
our method works for any SOGAT. For our particular SOGAT, the single
substitution syntax can be simplified, and we even have that any
single substitution model is equivalent to a parallel model in a weak
sense. This shows that if we have enough type formers, the
substitution calculus can be very weak, and still enough to bootstrap
the usual theory. It needs further investigation whether we can remove
even more structural rules, and still obtain isomorphic syntaxes. We
would also like to understand what a good notion of weak equivalence
is for single substitution models.
